\thispagestyle{plain}

\begin{center}
    \textbf{UCL Research Paper Declaration Form}
\end{center}

\textbf{Published Manuscripts}

\begin{enumerate}
    \item PyExaFMM: an exercise in designing high-performance software with Python and Numba.
    \begin{enumerate}[label=\alph*)]
      \item \href{https://ieeexplore.ieee.org/document/10124108}{DOI: 10.1109/MCSE.2023.3258288}
      \item Journal: Computing in Science \& Engineering
      \item Publisher: IEEE
      \item Date of Publication: Sept-Oct 2022
      \item Authors: Srinath Kailasa, Tingyu Wang, Lorena A. Barba, Timo Betcke
      \item Peer Reviewed: Yes
      \item Copyright Retained: No
      \item \href{https://doi.org/10.48550/arXiv.2303.08394}{ArXiv: 10.48550/arXiv.2303.08394}
      \item Associated Thesis Chapters: 1
      \item Statement of Contribution
      \begin{enumerate}
        \item Srinath Kailasa was the lead author and responsible for the direction of this research and the preparation of the manuscript.
        \item Tingyu Wang offered expert guidance as on fast multipole method implementations, and critical feedback of the manuscript.
        \item Lorena A. Barba served as an advisor, offering valuable insights into the structure of the manuscript, suggesting improvements to enhance clarity and impact of the work.
        \item Timo Betcke provided significant advisory support, contributing to the design and framing of the research, interpretation of the results and provided critical feedback of the manuscript.
      \end{enumerate}
    \end{enumerate}
\end{enumerate}

\textbf{Unpublished Manuscripts}

\begin{enumerate}
    \item M2L Translation Operators for Kernel Independent Fast Multipole Methods on Modern Architectures.
    \begin{enumerate}[label=\alph*)]
      \item \href{https://ieeexplore.ieee.org/document/10124108}{DOI: TODO}
      \item Intended Journal: SIAM Journal on Scientific Computing
      \item Authors: Srinath Kailasa, Timo Betcke, Sarah El-Kazdadi
      \item  \href{https://doi.org/10.48550/arXiv.2408.07436}{ArXiv: 10.48550/arXiv.2408.07436}
      \item Stage of Publication: Submitted
      \item Associated Thesis Chapters: 2
      \item Statement of Contribution
      \begin{enumerate}
        \item Srinath Kailasa was the lead author and responsible for the direction of this research and the preparation of the manuscript.
        \item Timo Betcke provided significant advisory support aid with interpretation of the results and provided critical feedback of the manuscript.
        \item Sarah El-Kazdadi provided expert guidance on the usage of explicit vector programming, which was critical for achieving the final results presented.
      \end{enumerate}
    \end{enumerate}

    \item kiFMM-rs: A Kernel-Independent Fast Multipole Framework in Rust
    \begin{enumerate}[label=\alph*)]
      \item \href{https://ieeexplore.ieee.org/document/10124108}{DOI: TODO}
      \item Intended Journal: Journal of Open Source Software
      \item Authors: Srinath Kailasa
      \item Stage of Publication: Submitted
      \item Associated Thesis Chapters: 1, 2
    \end{enumerate}

\end{enumerate}

\textbf{e-Signatures confirming that the information above is accurate}

\hspace*{10mm}

\textbf{Candidate:}

\textbf{Date:}

\hspace*{10mm}

\textbf{Supervisor/Senior Author:}

\textbf{Date:}